A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-10-25 LyapDraft.tex
making several tweaks to the intro based on some feedback

Momentum is a powerful heuristic for accelerating the convergence of optimization methods.    One can intuitively ``add momentum'' to a method by adding to the current step a weighted version of the previous step, encouraging the method to move along search directions that had been previously seen to be fruitful.  Such methods were first popularized by Polyak~\cite{Polyak1964}, and have been employed in many practical optimization solvers.  In particular, since the 1980s, momentum methods have been popular in neural networks as a way to accelerate the backpropagation algorithm.  Here, the intuition is that momentum allows local search to avoid ``long ravines'' and ``sharp curvatures'' in the sublevel sets of cost functions~\cite{Rumelhardt}.

Polyak motivated momentum methods by an analogy to a ``heavy ball'' moving in a potential well defined by the cost function.  However, Polyak's physical intuition was exceedingly difficult to make rigorous.  For quadratics costs, Polyak was able to apply an eigenvalue argument that showed that his Heavy Ball method required no more iterations than the method of conjugate gradients~\cite{Polyak1964}.  
Indeed, when applied to positive definite quadratic cost functions, Polyak's Heavy Ball method is equivalent to Chebyshev's iterative method~\cite{Chebyshev}.   Despite its intuitive elegance, Polyak's eigenvalue analysis does not apply globally for general convex functions. Indeed, Lessard~\emph{et al} derived a simple one dimensional counterexample where the standard Heavy-Ball method does not converge~\cite{Lessard14}.

In order to make momentum methods rigorous, a different approach was required.  In celebrated work, Nesterov devised a general scheme to accelerate convex optimization methods, which achieve optimal running times for a myriad of oracle models in convex programming~\cite{Nesterov04}.  To achieve such rigorous general applicability, however, Nesterov's proof techniques abandoned the physical intuition of Polyak~\cite{Nesterov04}.  In lieu of differential equations and potential functions, Nesterov devised the method of \emph{estimate sequences} to verify the correctness of his methods.  Researchers have struggled with understanding the intuition and underpinnings of the estimate sequence methodology since Nesterov's initial papers.  Nesterov himself often refers to the associated proof techniques as an ``algebraic trick.''

To overcome this lack of intuition, several authors have recently proposed schemes to achieve acceleration without appealing to estimate sequences~\cite{Fazel, BubeckLeeSingh15, Lessard14}.  Orthogonally, another set of authors have taken a different approach by analyzing the continuous-time limit of accelerated methods and showing that the stability of the resulting ODEs can be verified by analyzing a simple Lyapunov function~\cite{SuBoydCandes14,Krichene15,Acceleration}.  Unfortunately, none of these works provide a clear path of moving from a continuous time ODE to a discrete time optimization algorithm.  There are a vast number of ways to discretize ODEs, but not all of them give rise to convergent methods.  Indeed, for unconstrained optimization on Euclidean space, Polyak's Heavy Ball method and Nesterov's optimal method have the same continuous time limit.

@@

In order to have stable dynamics, the functions $\alpha_t$, $\beta_t$, and $\gamma_t$ must satisfy the necessary conditions
\begin{subequations}\label{Eq:IdeSca}
\begin{align}
\dot \beta_t \,&\leq\, e^{\alpha_t}   \label{Eq:IdeScaBet} \\
\dot \gamma_t \,&=\, e^{\alpha_t}  \label{Eq:IdeScaGam}.
\end{align}
\end{subequations}
These conditions were dubbed the~\emph{ideal scaling}, and they will play a predominant role in the present work.  Wibosono et al~\cite{Acceleration} also require that  $f$ and $h$ be differentiable and convex, and that $h$ be strictly convex. Under the ideal scaling, it was demonstrated that the momentum dynamic~\eqref{Eq:ELBreg} is the Euler-Lagrange equation (stationary condition) of a functional called the {\em Bregman Lagrangian}. This means that the family of accelerated gradient dynamics -- like the gradient flows which generate gradient methods -- can be understood as being generated from a {\em variational principal}. They also generalize the Lyapunov function of Su et al~\cite{SuBoydCandes14} and use it to show that the family of momentum dynamics~\eqref{Eq:ELBreg} obtain the convergence rate $O(e^{-\beta_t})$. However, they did not provide a Lyapunov analysis of the discrete-time algorithm, nor did they make explicit the connection between the Lyapunov function and the estimate sequence technique. We address some of these issues in the present work, providing a framework that generalizes these earlier results.

@@

%\begin{small}
\begin{subequations}\label{Eq:LyapAnal}
\begin{align}
 \frac{d}{dt}D_h\left(x, Z_t\right)  &= \frac{d}{dt}\left( h(x) - h(Z_t) - \langle \nabla h(Z_t), x- Z_t\rangle \right) \notag\\
& = - \langle \nabla h(Z_t), \dot Z_t\rangle -\left\langle \frac{d}{dt} \nabla h(Z_t),x - Z_t\right\rangle  + \langle \nabla h(Z_t), \dot Z_t\rangle \notag\\ 
 &= - \left\langle \frac{d}{dt} \nabla h\left(Z_t\right), x - Z_t\right\rangle \, \notag\\%\label{eq:apply-identity}\\
&=    \left\langle \frac{d}{dt}\left(e^{\beta_t}\right)\nabla f(X_t), x  - X_{t} - \frac{1}{\dot \beta_t} \dot X_{s} \right\rangle  \, \label{eq:apply-dynamics}   \\
& =  \frac{d}{dt}\left(e^{\beta_t}\right) \langle \nabla f(X_t), x - X_t \rangle\, - e^{ \beta_t} \langle \nabla f(X_t), \dot X_t\rangle dt\, \notag   \\
& = \frac{d}{dt} \left(e^{\beta_t}\right) \langle \nabla f(X_t), x - X_t \rangle\,  -  \frac{d}{dt} \left(e^{ \beta_t} f(X_t)\right)\,   + \frac{d}{dt} \left(e^{\beta_t}\right)  f(X_t)  \notag\\
&= \frac{d}{dt}\left( e^{\beta_t}\right) [f(X_t)+ \langle \nabla f(X_t), x - X_t \rangle] -\frac{d}{dt} \left(e^{ \beta_t} f(X_t)\right)\,  \notag  \\ %\label{Eq:Line9}\\
& \leq   \frac{d}{dt}\left( e^{\beta_t}\right) f(x) - \frac{d}{dt} \left(e^{ \beta_t} f(X_t)\right)\, \label{eq:apply-convexity}\\
& = - \frac{d}{dt} \left(e^{\beta_t}\left(f(X_t) - f(x)\right)\right) \label{Eq:cond1}.
\end{align}
\end{subequations}
%\end{small}
Here~\eqref{eq:apply-dynamics} uses the momentum dynamics~\eqref{Eq:ELH} and~\eqref{Eq:ELZ}. The inequality~\eqref{eq:apply-convexity} follows from the convexity of $f$. Rearranging terms and taking $x = x^\ast$, we have show that the function~\eqref{Eq:E} has nonpositive derivative for all $t$ and is hence a Lyapunov function for the dynamical system~\eqref{Eq:ELBreg}.

@@

however further exploration of this Lagrangian is outside the scope of this work. Here, we demonstrate that~\eqref{Eq:StrongLyap} can be used to show an $O(e^{-\beta_t})$ rate of convergence for~\eqref{Eq:StrongDyn} with a strong convexity assumption. To do so, note that if we can ensure  $\dot \E_t = e^{\beta_t}\dot \beta_t \tilde{\E}_t + e^{\beta_t} \dot{\tilde{\E}}_t \leq 0$, which amounts to ensuring

@@

\subsection{Searching for invariants}
Earlier work by Drori and Teboulle~\cite{DroriTeboulle13}, Kim and Fessler~\cite{Kim2016}, Taylor \emph{et al}~\cite{Taylor2016}, and Lessard~\emph{et al}~\cite{Lessard14} have shown that optimization algorithms could be analyzed by solving convex programming problems.  In particular, Lessard~\emph{et al} show that Lyapunov-like potential functions called \emph{integral quadratic constraints} can be found by solving a constant-sized semidefinite programming problem. It would be interesting to see if these results can be adapted to directly search for Lyapunov functions like those studied in this paper.  This would provide a method to automate the analysis of new techniques, possibly moving beyond momentum methods to novel families of optimization techniques.
\section*{Acknowledgement}
We would like to give special thanks to Andre Wibisono for the many helpful discussion involving this paper. We would also like to thank Orianna Demassi for several helpful suggestions and Stephen Tu who caught a small error in a previous version of this paper.
\bibliographystyle{plain}
\bibliography{refs.bib}

>>>>

Momentum is a powerful heuristic for accelerating the convergence of optimization methods.    One can intuitively ``add momentum'' to a method by adding to the current step a weighted version of the previous step, encouraging the method to move along search directions that had been previously seen to be fruitful.  Such methods were first popularized by Polyak~\cite{Polyak1964}, and have been employed in many practical optimization solvers.  In particular, since the 1980s, momentum methods have been popular in neural networks as a way to accelerate the backpropagation algorithm.  The intuition is that momentum allows local search to avoid ``long ravines'' and ``sharp curvatures'' in the sublevel sets of cost functions~\cite{Rumelhardt}.

Polyak motivated momentum methods by an analogy to a ``heavy ball'' moving in a potential well defined by the cost function.  However, Polyak's physical intuition was exceedingly difficult to make rigorous.  For quadratics costs, Polyak was able to apply an eigenvalue argument that showed that his Heavy Ball method required no more iterations than the method of conjugate gradients~\cite{Polyak1964}.  
Indeed, when applied to positive definite quadratic cost functions, Polyak's Heavy Ball method is equivalent to Chebyshev's Iterative Method~\cite{Chebyshev}.   Despite its intuitive elegance, Polyak's eigenvalue analysis does not apply globally for general convex functions. Indeed, Lessard~\emph{et al.} derived a simple one dimensional counterexample where the standard Heavy Ball method does not converge~\cite{Lessard14}.

In order to make momentum methods rigorous, a different approach was required.  In celebrated work, Nesterov devised a general scheme to accelerate convex optimization methods, which achieve optimal running times for a myriad of oracle models in convex programming~\cite{Nesterov04}. However, to achieve such rigorous general applicability, Nesterov's proof techniques abandoned the physical intuition of Polyak~\cite{Nesterov04}.  In lieu of differential equations and potential functions, Nesterov devised the method of \emph{estimate sequences} to verify the correctness of these momentum-based methods.  Researchers have struggled with understanding the intuition and underpinnings of the estimate sequence methodology since Nesterov's initial papers.  Nesterov himself often refers to the associated proof techniques as an ``algebraic trick.''

To overcome the lack of intuition for the estimate sequences technique, several authors have recently proposed schemes to achieve acceleration without appealing to estimate sequences~\cite{Fazel, BubeckLeeSingh15, Lessard14}.  Orthogonally, another set of authors have taken a different approach by analyzing the continuous-time limit of accelerated methods and showing that the stability of the resulting ODEs can be verified by analyzing a simple Lyapunov function~\cite{SuBoydCandes14,Krichene15,Acceleration}.  Unfortunately, none of these works provide a clear path of obtaining a discrete time optimization algorithm from a continuous time ODE.  There are a vast number of ways to discretize ODEs, but not all of them give rise to convergent methods.  Indeed, for unconstrained optimization on Euclidean space, Polyak's Heavy Ball method and Nesterov's accelerated gradient descent in the setting where the objective is strongly convex have the same continuous time limit.

@@

In order to have stable dynamics, the functions $\alpha_t$, $\beta_t$, and $\gamma_t$ must satisfy the necessary conditions
\begin{subequations}\label{Eq:IdeSca}
\begin{align}
\dot \beta_t \,&\leq\, e^{\alpha_t}   \label{Eq:IdeScaBet} \\
\dot \gamma_t \,&=\, e^{\alpha_t}  \label{Eq:IdeScaGam}.
\end{align}
\end{subequations}
These conditions were dubbed the~\emph{ideal scaling}, and they will play a predominant role in the present work.  Wibosono {\em et al.}~\cite{Acceleration} also require that  $f$ and $h$ be differentiable and convex, and that $h$ be strictly convex. Under the ideal scaling, it was demonstrated that the momentum dynamic~\eqref{Eq:ELBreg} is the Euler-Lagrange equation (stationary condition) of a functional called the {\em Bregman Lagrangian}. This means that the family of accelerated gradient dynamics -- like the gradient flows which generate gradient methods -- can be understood as being generated from a {\em variational principal}. They also generalize the Lyapunov function of Su {\em et al.}~\cite{SuBoydCandes14} and use it to show that the family of momentum dynamics~\eqref{Eq:ELBreg} obtain the convergence rate $O(e^{-\beta_t})$. However, they did not provide a Lyapunov analysis of the discrete-time algorithm, nor did they make explicit the connection between the Lyapunov function and the estimate sequence technique. We address some of these issues in the present work, providing a framework that generalizes these earlier results.

@@

%\begin{small}
\begin{subequations}\label{Eq:LyapAnal}
\begin{align}
 \frac{d}{dt}D_h\left(x, Z_t\right)  &= \frac{d}{dt}\left( h(x) - h(Z_t) - \langle \nabla h(Z_t), x- Z_t\rangle \right) \notag\\
& = - \langle \nabla h(Z_t), \dot Z_t\rangle -\left\langle \frac{d}{dt} \nabla h(Z_t),x - Z_t\right\rangle  + \langle \nabla h(Z_t), \dot Z_t\rangle \notag\\ 
 &= - \left\langle \frac{d}{dt} \nabla h\left(Z_t\right), x - Z_t\right\rangle \, \notag\\%\label{eq:apply-identity}\\
&=    \left\langle \frac{d}{dt}\left(e^{\beta_t}\right)\nabla f(X_t), x  - X_{t} - \frac{1}{\dot \beta_t} \dot X_{s} \right\rangle  \, \label{eq:apply-dynamics}   \\
& =  \frac{d}{dt}\left(e^{\beta_t}\right) \langle \nabla f(X_t), x - X_t \rangle\, - e^{ \beta_t} \langle \nabla f(X_t), \dot X_t\rangle dt\, \notag   \\
& = \frac{d}{dt} \left(e^{\beta_t}\right) \langle \nabla f(X_t), x - X_t \rangle\,  -  \frac{d}{dt} \left(e^{ \beta_t} f(X_t)\right)\,   + \frac{d}{dt} \left(e^{\beta_t}\right)  f(X_t)  \notag\\
&= \frac{d}{dt}\left( e^{\beta_t}\right) [f(X_t)+ \langle \nabla f(X_t), x - X_t \rangle] -\frac{d}{dt} \left(e^{ \beta_t} f(X_t)\right)\,  \notag  \\ %\label{Eq:Line9}\\
& \leq   \frac{d}{dt}\left( e^{\beta_t}\right) f(x) - \frac{d}{dt} \left(e^{ \beta_t} f(X_t)\right)\, \label{eq:apply-convexity}\\
& = - \frac{d}{dt} \left(e^{\beta_t}\left(f(X_t) - f(x)\right)\right) \label{Eq:cond1}.
\end{align}
\end{subequations}
%\end{small}
Here~\eqref{eq:apply-dynamics} uses the momentum dynamics~\eqref{Eq:ELH} and~\eqref{Eq:ELZ}. The inequality~\eqref{eq:apply-convexity} follows from the convexity of $f$. Rearranging terms and taking $x = x^\ast$, we have show that the function~\eqref{Eq:E} has nonpositive derivative for all $t$ and is hence a Lyapunov function for the dynamical system~\eqref{Eq:EL}.

@@

and that this Lagrangian can further generalized; however, further exploration of this Lagrangian is outside the scope of this work. Here, we demonstrate that~\eqref{Eq:StrongLyap} can be used to show an $O(e^{-\beta_t})$ rate of convergence for~\eqref{Eq:StrongDyn} with a strong convexity assumption. To do so, note that if we can ensure  $\dot \E_t = e^{\beta_t}\dot \beta_t \tilde{\E}_t + e^{\beta_t} \dot{\tilde{\E}}_t \leq 0$, which amounts to ensuring

@@

\subsection{Searching for invariants}
Earlier work by Drori and Teboulle~\cite{DroriTeboulle13}, Kim and Fessler~\cite{Kim2016}, Taylor \emph{et al}~\cite{Taylor2016}, and Lessard~\emph{et al}~\cite{Lessard14} have shown that optimization algorithms could be analyzed by solving convex programming problems.  In particular, Lessard~\emph{et al} show that Lyapunov-like potential functions called \emph{integral quadratic constraints} can be found by solving a constant-sized semidefinite programming problem. It would be interesting to see if these results can be adapted to directly search for Lyapunov functions like those studied in this paper.  This would provide a method to automate the analysis of new techniques, possibly moving beyond momentum methods to novel families of optimization techniques.
\section*{Acknowledgements}
We would like to give special thanks to Andre Wibisono for the many helpful discussion involving this paper. We would also like to thank Orianna Demassi for several helpful suggestions and Stephen Tu who caught a small error in a previous version of this paper.
\bibliographystyle{plain}
\bibliography{refs.bib}
